D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-03 ootavacuous.tex
Edits on top of the recent batch of changes
These edits refine the changes committed in the last day or two.  They
also remove the waffling on the term "loose C++", nobody seems to mind
it any more.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above.
{\it Contra\/} that article, we will use the term \emph{loose C++}\footnote{
	There is some discomfort with this name, so this footnote
	is reserved for future bikeshedding.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

The first line identifies it as a C-language litmus test and gives it
a name, and this name identifies the litmus test's source file within
the \path{litmus} directory in the \url{https://github.com/paulmckrcu/oota}
repository.
Lines~2--5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Variables that are not explicitly initialized are initialized by
default to zero.
Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
define the second thread, \co{P1()}.\footnote{
	The ``P'' stands for ``process'', which is \co{herd7}'s name
	for ``thread''.}
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each thread may access, in this case, \co{x}
and \co{y}.
The body of each thread contains a limited subset of C++ code
(written in a slightly stilted manner to keep the load operations separate
from the rest).

@@

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the loose C++
memory model by assuming each thread's load reads from the other's
store.\footnote{
	Such a compiler could presumably instead choose to avoid this
	OOTA cycle, for example, by retaining the loads from \co{x}
	and \co{y}.}
A similar justification can underlie the reasoning in
Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.

@@

This paper shows that semantic dependencies are relative to specific
executions of a program rather than strictly of the source code of that
program.
Future work might expand on
Section~\ref{sec:Relative versus Absolute Dependency},
further exploring absolute semantic dependencies that are special cases
where semantic dependencies can be a function strictly of the source code.

This paper considers only fragments of programs whose threads
communicate using shared memory.
Future work might include the effects of input and output, as well as
code that reads out the current time, that uses thread IDs, or that
operates on pointer values
(see Section~\ref{sec:Effect of Memory Layout}).

@@

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application of the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This refinement, combined with a survey of code containing non-trivial
semantic dependencies, uncovered some shortcomings in typical definitions of
``semantic dependency'', pointing out examples where semantic dependencies
are a function of an execution rather than of the source code, and also
examples where a single semantic dependency extends from multiple loads
to a store, but not from any one of those loads to that same store.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many To One}.

This generalization, combined with a focus on compiler-based C++
implementations using single-threaded analyses, led to the fundamental
property of semantic dependency shown in
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
This property in turn led to the precise definition of ``semantic
dependency'' shown in
Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
These definitions are then used in
Section~\ref{sec:Verifying the Fundamental Property}
to demonstrate that correct real-world compiler-based C++
implementations relying on single-threaded analyses are inherently
incapable of producing OOTA cycles in UB-free programs using volatile
atomics.

C++ implementations that additionally confine optimizations of code
containing nonvolatile atomic operations to the quasi-volatile limitations
outlined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
are incapable of producing OOTA cycles in UB-free programs using
nonvolatile as well as volatile atomics.

>>>>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above.
{\it Contra\/} that article, we will use the term \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

The first line identifies it as a C-language litmus test and gives it
a name, and this name identifies the litmus test's source file within
the \path{litmus} directory in the \url{https://github.com/paulmckrcu/oota}
repository.
Lines~2--5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Variables that are not explicitly initialized are initialized
to zero by default.
Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
define the second thread, \co{P1()}.\footnote{
	The ``P'' stands for ``process'', which is \co{herd7}'s name
	for ``thread''.}
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each thread may access, in this case, \co{x}
and \co{y}.
The body of each thread contains a limited subset of C++ code
(written in a slightly stilted manner to keep the load operations separate
from the rest).

@@

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the loose C++
memory model by assuming each thread's load reads from the other's
store.\footnote{
	A less perverse compiler could choose to avoid the
	OOTA cycle simply by not making this transformation.}
A similar justification can underlie the reasoning in
Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.

@@

This paper shows that semantic dependencies are relative to specific
executions of a program and the specific compiler in use rather than
being determined solely by the source code.
Future work might expand on
Section~\ref{sec:Relative versus Absolute Dependency},
exploring special cases in which absolute semantic dependencies are
functions just of the source code.

This paper considers only programs whose threads
communicate using shared memory.
Future work might include the effects of input and output or
other operations which can vary from one execution to another,
such as those based on the current time, process IDs, or pointer values
(see Section~\ref{sec:Effect of Memory Layout}).

@@

This paper has presented samples of code containing non-trivial
semantic dependencies, uncovering some shortcomings in typical definitions of
``semantic dependency''.
It pointed out examples where semantic dependencies
are a function of an execution rather than of the source code, and examples
where a single semantic dependency extends from multiple loads
to a store but not from any one of those loads.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many To One}.

This generalization, combined with a focus on compiler-based loose C++
implementations using single-thread analysis, and a consideration of
the relation between source code and machine code, led to the formulation
of the Fundamental Property of semantic dependency shown in
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
This in turn led to a precise definition of ``semantic dependency'' given in
Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
This definition was used in
Section~\ref{sec:Verifying the Fundamental Property}
to demonstrate that these implementations are
incapable of producing OOTA cycles in UB-free programs,
provided they treat all nonvolatile atomic objects as
quasi volatile, obeying the limitations outlined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

A variant definition of ``semantic dependency'' given in
Section~\ref{sec:For Compilers Using Global Analysis}, appropriate for
compiler-based loose C++ implementations that don't restrict
their code transformations to those based on single-thread analysis,
was used to show that such implementations
are incapable of producing OOTA cycles in UB-free programs
provided they treat all atomic objects as volatile.
